---
title: "Chapter 12 - Exercises"
output: html_notebook
---

```{r include=FALSE}
library(here)
library(rjags)
library(ggplot2)
library(reshape2)
library(dplyr)
theme_set(theme_classic())
```

## Exercise 1

Specification of fixed- and random-effects in WinBUGS: The WinBUGS model
description for the random-intercepts, random-slope model (i.e., the second
one we fit in this chapter) is very similar to the fixed-effects “version” of the
same model, i.e., the one we fitted in Chapter 11. Without looking at the
WinBUGS model description in that chapter, take the linear mixed model
description for WinBUGS from the current chapter and change it back to a
fixed-effects ANCOVA with population-specific intercepts and slopes, i.e.,
corresponding to what you would fit in R as lm(mass ~ pop * length).

```{r}
n.groups <- 56				# Number of populations
n.sample <- 10				# Number of vipers in each pop
n <- n.groups * n.sample 		# Total number of data points
pop <- gl(n = n.groups, k = n.sample) 	# Indicator for population
original.length <- runif(n, 45, 70) 
length <- scale(original.length)

Xmat <- model.matrix(~pop*length-1-length)

intercept.mean <- 230			# mu_alpha
intercept.sd <- 20			# sigma_alpha
slope.mean <- 60			# mu_beta
slope.sd <- 30				# sigma_beta

intercept.effects <- rnorm(n = n.groups, mean = intercept.mean, sd = intercept.sd)
slope.effects <- rnorm(n = n.groups, mean = slope.mean, sd = slope.sd)
all.effects <- c(intercept.effects, slope.effects) # Put them all together

lin.pred <- Xmat[,] %*% all.effects	# Value of lin.predictor
eps <- rnorm(n = n, mean = 0, sd = 30)	# residuals 
mass <- lin.pred + eps
```


# Define model

Instead of this snippet for the Prior used in the random effects model:

```{r}
# Priors
 for (i in 1:ngroups){		
    alpha[i] ~ dnorm(mu.int, tau.int)	# Random intercepts
    beta[i] ~ dnorm(mu.slope, tau.slope)# Random slopes
 }
```

We replace it with fixed effects estimation of slope and intercept.

```{r}
model <- "
model {

# Priors
 for (i in 1:ngroups){		
    alpha[i] ~ dnorm(0, 0.00001)
    beta[i] ~ dnorm(0, 0.00001)
 }

 tau <- 1 / ( sigma * sigma)		# Residual precision
 sigma ~ dunif(0, 100)			# Residual standard deviation

# Likelihood
 for (i in 1:n) {
    mass[i] ~ dnorm(mu[i], tau)
    mu[i] <- alpha[pop[i]] + beta[pop[i]]* length[i]
 }

 slope = sum(beta)/ngroups
 intercept = sum(alpha)/ngroups
 slope.sd = sd(beta)
 intercept.sd = sd(alpha)
}
"
```

```{r include=FALSE}
data <- list("length" = as.numeric(length),
             "mass" = as.numeric(mass),
             "pop" = as.numeric(pop),
             "n" = n,
             "ngroups" = max(as.numeric(pop)))

inits <- function(){ list(alpha = rnorm(n.groups, 0, 2), 
                          beta = rnorm(n.groups, 1, 1),
                          sigma = rlnorm(1))}

params <- c("slope", "slope.sd", "intercept", "intercept.sd", "sigma")

jagsModel <- jags.model(file = textConnection(model),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 500)

Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 2000,
                        thin = 2)
```

```{r}
summary(Samples)
```

## Exercise 2

Swiss haredata: Fit a random-coefficients regression without intercept–slope
correlation to mean density(i.e., ~ population * year, with year
continuous).

```{r}
hares <- read.table(file = here("data","hares.data.txt"), header = TRUE)

hares. <- hares %>%
  filter(!is.na(mean.density))
hares.$scaled.year <- scale(hares.$year)
```

```{r}
model <- "
model {

# Priors
 for (i in 1:ngroups){		
    alpha[i] ~ dnorm(mu.int, tau.int)	# Random intercepts
    beta[i] ~ dnorm(mu.slope, tau.slope)# Random slopes
 }

 mu.int ~ dnorm(0, 0.001)		# Mean hyperparameter for random intercepts
 tau.int <- 1 / (sigma.int * sigma.int)
 sigma.int ~ dunif(0, 100)		# SD hyperparameter for random intercepts

 mu.slope ~ dnorm(0, 0.001)		# Mean hyperparameter for random slopes
 tau.slope <- 1 / (sigma.slope * sigma.slope)
 sigma.slope ~ dunif(0, 100)		# SD hyperparameter for slopes

 tau <- 1 / ( sigma * sigma)		# Residual precision
 sigma ~ dunif(0, 100)			# Residual standard deviation

# Likelihood
 for (i in 1:n) {
    mean.density[i] ~ dnorm(mu[i], tau)
    mu[i] <- alpha[site[i]] + beta[site[i]]* year[i]
 }
}
"
```

```{r include=FALSE}
data <- list("mean.density" = hares.$mean.density,
             "year" = as.numeric(hares.$scaled.year),
             "site" = as.numeric(hares.$site),
             "n" = nrow(hares.),
             "ngroups" = max(as.numeric(hares.$site)))

inits <- function(){ list(alpha = rnorm(max(as.numeric(hares.$site)), 0, 2), 
                          beta = rnorm(max(as.numeric(hares.$site)), 0, 2),
                          mu.int = rnorm(1, 0, 1),
                          mu.slope = rnorm(1, 0, 1),
                          sigma.int = rlnorm(1),
                          sigma = rlnorm(1),
                          sigma.slope = rlnorm(1))}

params <- c( "mu.int", "mu.slope", "sigma.int", "sigma.slope", "sigma")

jagsModel <- jags.model(file = textConnection(model),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 500)

Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000,
                        thin = 2)
```

```{r}
summary(Samples)
```

