---
title: "Chapter 11 - Exercises"
output: html_notebook
---

```{r include=FALSE}
library(here)
library(rjags)
library(ggplot2)
library(dplyr)
library(reshape2)
theme_set(theme_classic())
```

## Exercise 1

1. Probability of a parameter: What is the probability that the slope of the mass – length relationship of asp vipers is inferior in the Jura than in the Massif Central? Produce a graphical and a numerical answer.

First load the data

```{r}
load(here("data", "vipers.rda"))
length.scaled <- scale(length)
```

Bundle data and run MCMC

```{r, include=FALSE}
data <- list("length" = as.numeric(length.scaled),
             "mass" = as.numeric(mass),
             "pop" = pop,
             "n" = n,
             "n.group" = n.groups)

inits <- function(){ list(alpha = runif(3, -1000, 1000), 
                          beta = runif(3, -1000, 1000),
                          sigma = runif(1, 0, 100) )}

params <- c("alpha", "beta", "sigma")

jagsModel <- jags.model(file = here("models", "ancova.txt"),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 500)

Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000)
```

```{r}
autocorr.plot(Samples[[1]][,1])
```
```{r}
effectiveSize(Samples)
```


```{r}
mcmc <- as.data.frame(do.call(rbind, Samples))
colnames(mcmc) <- c("a.Pyrenees", "a.Massif", "a.Jura", "b.Pyrenees", "b.Massif", "b.Jura", "sigma")

mcmc$diff.b.Mas.Jur <- mcmc$b.Massif - mcmc$b.Jura

p.data <- melt(mcmc)
sum(mcmc$diff.b.Mas.Jur > 0) / length(mcmc$diff.b.Mas.Jur)
```
There is a 85% chance that the vipers in Massif Central are bigger than in Jura.

```{r}
ggplot(data = subset(p.data, variable == "diff.b.Mas.Jur"), aes(x = value)) + 
  geom_histogram(aes(y = ..density..), colour = "black", fill = "white", binwidth = 1) +
  geom_density(alpha = .2, fill = "#FF6666") +
  geom_vline(xintercept = 0, size = 2, colour = "red")
```
The difference between Massif Central and Jura vipers.
```{r}
ggplot(data = subset(p.data, variable == "b.Massif" | variable == "b.Jura"), aes(x = value, color = variable, fill = variable)) + 
  geom_histogram(aes(y = ..density..), alpha = .5, binwidth = 1, position = "identity") +
  geom_density(alpha = .2)
  #geom_vline(xintercept = 0, size = 2, colour = "red")
```
Both distribution for slope of the Massif and Jura vipers plotted separately.


## Exercise 2

2. Related models: Adapt the code to fit two variations of the model:
• Fit different intercepts but a common slope
• Fit the same intercept and the same slope

### Model with common slope

```{r}
model.common.slope <- "
model {
# Likelihood
 for (i in 1:n) {
    mass[i] ~ dnorm(mu[i], tau)
    mu[i] <- alpha[pop[i]] + beta * length[i]
 }

# Priors
 for (i in 1:n.group){		
    alpha[i] ~ dunif(-1000, 1000)		# Intercepts
 }
 beta ~ dunif(-1000, 1000)		# Slope
 sigma ~ dunif(0, 100)			# Residual standard deviation
 tau <- 1 / ( sigma * sigma)
}
"
```

```{r}
data <- list("length" = as.numeric(length.scaled),
             "mass" = as.numeric(mass),
             "pop" = pop,
             "n" = n,
             "n.group" = n.groups)

inits <- function(){ list(alpha = runif(3, -1000, 1000), 
                          beta = runif(1, -1000, 1000),
                          sigma = runif(1, 0, 100) )}

params <- c("alpha", "beta", "sigma")

jags.common.slope <- jags.model(file = textConnection(model.common.slope),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 500)

common.slope <- coda.samples(jags.common.slope, 
                        variable.names = params, 
                        n.iter = 1000)
```

### Model shared Intercept and Slope

```{r}
model.shared <- "
model {
# Likelihood
 for (i in 1:n) {
    mass[i] ~ dnorm(mu[i], tau)
    mu[i] <- alpha + beta * length[i]
 }

# Priors
 alpha ~ dunif(-1000, 1000)		# Intercept
 beta ~ dunif(-1000, 1000)		# Slope
 sigma ~ dunif(0, 100)			# Residual standard deviation
 tau <- 1 / ( sigma * sigma)
}
"
```

```{r}

inits <- function(){ list(alpha = runif(1, -1000, 1000), 
                          beta = runif(1, -1000, 1000),
                          sigma = runif(1, 0, 100) )}

jags.shared <- jags.model(file = textConnection(model.shared),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 500)

shared.intercept <- coda.samples(jags.shared, 
                        variable.names = params, 
                        n.iter = 1000)
```

```{r}
c.slope <- as.data.frame(do.call(rbind, common.slope))
means <- summarise_all(c.slope, funs(mean))

shared <- as.data.frame(do.call(rbind, shared.intercept))
shared.means <- summarise_all(shared, funs(mean))

p.data <- data.frame(mass = mass,
                     length = length.scaled,
                     pop = pop)
c.slope.lines <- data.frame(intercept = as.numeric(means[,1:3]),
                            slope = rep(means[,4], 3),
                            pop = levels(pop))

ggplot(data = p.data) + 
  geom_point(aes(y = mass, x = length, color = pop)) + 
  geom_abline(data = c.slope.lines, aes(slope = slope, intercept = intercept, color = pop)) +
  geom_abline(data = shared.means, aes(slope = beta, intercept = alpha))
```
All estimated lines are plotted. Black line is the shared slope and intercept.

## Exercise 3

3. Quadratic effects: Add a quadratic term to the mass – length relationship,
i.e., fit the model pop + (length + length^2). You do not need to
reassemble a dataset that contains an effect of length squared, but you can
simply take the dataset we have already created in this chapter.

```{r}
model <- "
model {
# Likelihood
 for (i in 1:n) {
    mass[i] ~ dnorm(mu[i], tau)
    mu[i] <- alpha[pop[i]] + beta[pop[i]] * length[i] + gamma[pop[i]] * length[i] * length[i]
 }

# Priors
 for (i in 1:n.group){		
    alpha[i] ~ dunif(-1000, 1000)
    beta[i] ~ dunif(-1000, 1000)
    gamma[i] ~ dunif(-1000, 1000)
 }
 sigma ~ dunif(0, 100)			# Residual standard deviation
 tau <- 1 / ( sigma * sigma)
 }
"
```

alternatively you can use pow(length, 2)

```{r}
data <- list("length" = as.numeric(length.scaled),
             "mass" = as.numeric(mass),
             "pop" = pop,
             "n" = n,
             "n.group" = n.groups)

inits <- function(){ list(alpha = runif(3, -1000, 1000), 
                          beta = runif(3, -1000, 1000),
                          gamma = runif(3, -1000, 1000),
                          sigma = runif(1, 0, 100))}

params <- c("alpha", "beta", "gamma", "sigma")

jagsModel <- jags.model(file = textConnection(model),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 500)

Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000)
```

```{r}
mcmc <- as.data.frame(do.call(rbind, Samples))
means <- as.numeric(summarise_all(mcmc, funs(mean)))

p.data <- data.frame(mass = mass,
                     length = length.scaled,
                     pop = pop)

ggplot(data = p.data) + 
  geom_point(aes(y = mass, x = length, color = pop)) +
  stat_function(colour = "red", fun = function(x) means[1] + means[4] * x + means[7] * x^2) +
  stat_function(colour = "green", fun = function(x) means[2] + means[5] * x + means[8] * x^2) +
  stat_function(colour = "blue", fun = function(x) means[3] + means[6] * x + means[9] * x^2)
```

4. Swiss haredata: Fit an ANCOVA (pop * year, with year as a continuous
explanatory variable) to the meandensity. Also compute residuals and plot
them to check for outliers.

```{r}
hares <- read.table(file = here("data","hares.data.txt"), header = TRUE)

hares. <- hares %>%
  filter(!is.na(mean.density))
hares.$scaled.year <- scale(hares.$year)
```

```{r}
model <- "
model {
# Likelihood
 for (i in 1:n) {
    mean.density[i] ~ dnorm(mu[i], tau)
    mu[i] <- alpha[site[i]] + beta[site[i]] * year[i]
    residuals[i] <- mean.density[i] - mu[i]
 }

# Priors
 for (i in 1:n.group){		
    alpha[i] ~ dunif(-1000, 1000)
    beta[i] ~ dunif(-1000, 1000)
 }
 sigma ~ dunif(0, 100)
 tau <- 1 / ( sigma * sigma)
 }
"
```

```{r}
data <- list("mean.density" = hares.$mean.density,
             "year" = as.numeric(hares.$scaled.year),
             "site" = hares.$site,
             "n" = nrow(hares.),
             "n.group" = length(levels(hares.$site)))

inits <- function(){ list(alpha = runif(56, -1000, 1000), 
                          beta = runif(56, -1000, 1000),
                          sigma = runif(1, 0, 100))}

params <- c("alpha", "beta", "sigma", "residuals")

jagsModel <- jags.model(file = textConnection(model),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 500)

Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000)
```

```{r}
mcmc <- as.data.frame(do.call(rbind, Samples))
means <- as.numeric(summarise_all(mcmc, funs(mean)))
means <- means[-c(1:112, 678)]
p.data <- melt(means)
p.data$x <- c(1:677)

ggplot(data = p.data, aes(x = x, y = value)) + geom_point()
#ggplot(data = hares.) +
#  geom_point(aes(x = scaled.year, y = mean.density, color = region))
```

