---
title: "Chapter 11 - Exercises"
output: html_notebook
---

```{r, include=FALSE}
library(here)
library(rjags)
library(ggplot2)
library(dplyr)
library(reshape2)
theme_set(theme_classic())
```

## Exercise 1

1. Probability of a parameter: What is the probability that the slope of the mass – length relationship of asp vipers is inferior in the Jura than in the Massif Central? Produce a graphical and a numerical answer.

First load the data

```{r}
load(here("data", "vipers.rda"))
length.scaled <- scale(length)
```

Bundle data and run MCMC

```{r, include=FALSE}
data <- list("length" = as.numeric(length.scaled),
             "mass" = as.numeric(mass),
             "pop" = pop,
             "n" = n,
             "n.group" = n.groups)

inits <- function(){ list(alpha = runif(3, -1000, 1000), 
                          beta = runif(3, -1000, 1000),
                          sigma = runif(1, 0, 100) )}

params <- c("alpha", "beta", "sigma")

jagsModel <- jags.model(file = here("models", "ancova.txt"),
                        data = data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 100)

Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000)
```



2. Related models: Adapt the code to fit two variations of the model:
• Fit different intercepts but a common slope
• Fit the same intercept and the same slope

3. Quadratic effects: Add a quadratic term to the mass – length relationship,
i.e., fit the model pop + (length + length^2). You do not need to
reassemble a dataset that contains an effect of length squared, but you can
simply take the dataset we have already created in this chapter.

4. Swiss haredata: Fit an ANCOVA (pop * year, with year as a continuous
explanatory variable) to the meandensity. Also compute residuals and plot
them to check for outliers.