---
title: "Chapter 5 - Exercises"
output:
  html_document: default
  html_notebook: default
---

Load required Packages for the Exercises and set a seed for reproducability 

```{r message=FALSE, error=FALSE}
library(here)
library(rjags)
library(ggplot2)
theme_set(theme_classic())
set.seed(101)

```

Write the model of the mean into a file and generate two datasets.

```{r}
# Generate two samples of body mass measurements of male peregrines
y10 <- rnorm(n = 10, mean = 600, sd = 30) # Sample of 10 birds
y1000 <- rnorm(n = 1000, mean = 600, sd = 30) # Sample of 1000 birds

model<-"
    model {
    # Likelihood
    for(i in 1:nobs){
    mass[i] ~ dnorm(population.mean, precision) # Normal parameterized by precision
    }

    # Priors
    population.mean ~ dunif(0,1000)		
    precision <- 1 / population.variance	# Precision = 1/variance
    population.variance <- population.sd * population.sd
    population.sd ~ dunif(0,100)
    }
    "
model.spec <- textConnection(model)
```

2. Exercise 
Run the model with the small and large dataset and explain the differences.
The model assumes wide uninformed priors for both mean and standard deviation.
We bundle and run the model two times with both datasets. The init and params stay the same for both models.
```{r}
data1 <- list(mass = y1000, nobs = length(y1000))
data2 <- list(mass = y10, nobs = length(y10))

inits <- list(population.mean = rnorm(1,550), 
              population.sd = runif(1, 1, 30))

params <- c("population.mean", 
            "population.sd", 
            "population.variance")

model.spec <- textConnection(model)
jagsModel1 <- jags.model(file= model.spec,
                        data=data1,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 100)

model.spec <- textConnection(model)
jagsModel2 <- jags.model(file= model.spec,
                        data=data2,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 100)

Samples1 <- coda.samples(jagsModel1, 
                        variable.names = params, 
                        n.iter = 1000)

Samples2 <- coda.samples(jagsModel2, 
                        variable.names = params, 
                        n.iter = 1000)
```

Plot the results for the large dataset. n = 1000

```{r}
plot(Samples1)
```

Plot the results for the small dataset. n = 10

```{r}
plot(Samples2)
```

The estimate for standard deviation and variance are skewed, mainly caused by the low sample size.
The model tries to accomodate the the data, but since it is skewed the marginal distribution will also be skewed.

Next we plot the data again. The means are different, with the large sample being closer to the true mean of the data generator (600).

```{r}
xlim = c(450, 750)
par(mfrow = c(2,1))
hist(y10, col = 'grey ', xlim = xlim, 
     main = paste('Body mass (g) of 10 male peregrines; mean:',
                  round(mean(do.call(rbind,Samples2)[,1]),1)))
hist(y1000, col = 'grey', xlim = xlim, 
     main = paste('Body mass (g) of 1000 male peregrines; mean:',
                  round(mean(do.call(rbind,Samples1)[,1]),1)))
```

5. Derived quantities

The Bayesian framework and the Bugs model language is very flexible in handling derived calculations. 
We can specify any kind of formula or variable that we want to know about and estimate it during sampling.
In the following model we added a line of code to calculate the coefficient of variance.
The following lines of code will fit the model to the large dataset like we did in the previous exercises.

```{r}
model<-"
    model {
# Likelihood
for(i in 1:nobs){
mass[i] ~ dnorm(population.mean, precision) # Normal parameterized by precision
}

# Priors
population.mean ~ dunif(0,1000)		
precision <- 1 / population.variance	# Precision = 1/variance
population.variance <- population.sd * population.sd
population.sd ~ dunif(0,100)

# Derived quantities
CV <- population.sd/population.mean
}
"
model.spec <- textConnection(model)

data <- list(mass = y1000, nobs = length(y1000))

inits <- function() list(population.mean = runif(1,0,1000), population.sd = runif(1, 1, 100))

params <- c("population.mean", 
            "population.sd", 
            "population.variance",
            "CV")

jagsModel <- jags.model(file= model.spec,
                        data=data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 200)
Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000)
```

The estimates CV is below with a 95% confidence interval. We are 95% sure that the CV for our falcon population is between 0.047 and 0.05.

```{r}
CV_estimate <- do.call(rbind,Samples)[,1]

cat("CV mean:", round(mean(CV_estimate),3),
    "\nConfidence interval\n    5%  95%\n", round(quantile(CV_estimate, probs = c(.05,.95)), 3), 
    "\n")
```

We can plot the posterior distribution.

```{r}
CV <- data.frame(CV = CV_estimate)
ggplot(data = CV, aes(x = CV))+ geom_density(fill = "grey80")
```

6. Fit the model of the mean to the hares data set. report the mean, SE of mean and a 95% CI for the mean hare density
First we load the dataset and filter it for empty entries.

```{r}
hares <- read.table(file = here("data","hares.data.txt"), header = TRUE)
hares <- hares[!is.na(hares$mean.density),]
```

We can use the same model as before.
Afterwards as before the data is bundled, the model initiated and estimates are drawn from the posterior.

```{r}
model.spec <- textConnection(model)

data <- list(mass = hares$mean.density, nobs = nrow(hares))

inits <- function() list(population.mean = runif(1, 0, 1000), population.sd = runif(1, 0, 100))

params <- c("population.mean", 
            "population.sd", 
            "population.variance",
            "CV")

jagsModel <- jags.model(file= model.spec,
                        data=data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 100)
Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000)
```

The mean estimate of the hare density and standard error is below, with confidence interval for the mean.

```{r}
hare_means <- data.frame(density = do.call(rbind,Samples)[,2], sd = do.call(rbind,Samples)[,3])

cat("hare mean density:", round(mean(hare_means$density),2),"\nSE:", round(mean(hare_means$sd),2),
    "\n\nConfidence interval of the mean\n    5%  95%\n", round(quantile(hare_means$density, probs = c(.05,.95)), 3)) 
```

Marginal estimate for the mean hare density

```{r}
ggplot(data = hare_means, aes(x = density))+ geom_density(fill = "grey80")
```

