---
title: "Chapter 5 - Book Code"
output:
  html_document: default
  html_notebook: default
---
Load required Packages for this Chapter and set a seed for reproducability

```{r message = FALSE}
require(here)
require(rjags)
set.seed(101)
```

5.2 Data Generation

Generate two samples of body mass measurements of male peregrines and plot them.

```{r}
y10 <- rnorm(n = 10, mean = 600, sd = 30) # Sample of 10 birds
y1000 <- rnorm(n = 1000, mean = 600, sd = 30) # Sample of 1000 birds

xlim = c(450, 750)
par(mfrow = c(2,1))
hist(y10, col = 'grey ', xlim = xlim, main = 'Body mass (g) of 10 male peregrines')
hist(y1000, col = 'grey', xlim = xlim, main = ' Body mass (g) of 1000 male peregrines')
```

Next we fit a simple linear model of the mean to the data of 1000 peregrine falcons.
The estimated mean is 607, slightly off from our value given in the data generation in rnorm()

```{r}
summary(lm(y10 ~ 1))
```

5.4 Model of the mean in RJAGS

We fit the model of the mean in the Bayes way with rjags.
First the model description is saved.
The normal distribution is specified by precision instead of standardvariation.
We specify the priors as flat priors. For the mean, we assume prior knowledge of a possible mean between 500 and 700 g of weight. 

```{r}
model<-"
    model {
    # Likelihood
    for(i in 1:nobs){
    mass[i] ~ dnorm(population.mean, precision)
    }

    # Priors
    population.mean ~ dunif(500,700)		
    precision <- 1 / population.variance
    population.variance <- population.sd * population.sd
    population.sd ~ dunif(0,100)
    }
    "
model.spec <- textConnection(model)
```

Next, the data is bundled, the starting values are initiated and the parameters that we want to observe (estimate) are saved in params. 
The last two commands initialize the model and the description is checked for inconsistencies. Furthermore the burn-in phase of the Markov Chain is run and the parameters were estimated.
The coda.samples() command draws the actual samples from the posterior.

```{r message = FALSE}
# Bundle data
data <- list(mass = y1000, nobs = length(y1000))

# Function to generate starting values
inits <- function(){
  list(population.mean = runif(1,500,700), population.sd = runif(1, 1, 30))}

# Parameters
params <- c("population.mean", "population.sd", "population.variance")

jagsModel <- jags.model(file= model.spec,
                        data=data,
                        init = inits,
                        n.chains = 3, 
                        n.adapt = 100)
Samples <- coda.samples(jagsModel, 
                        variable.names = params, 
                        n.iter = 1000)
```

Lastly we plot the results from the Markov-Chain and check for Convergence.

```{r}
plot(Samples)
```

Here are the outputs from the analyses, where we can see the mean and standard deviation that we estimated using the Bayesian model of the mean.
We have mean and standard deviation for our estimates as well as quantiles for each variable.

```{r}
summary(Samples)
```

